---
title: "Hands-on_Ex10"
author: "Lin Yan"
date: "21 Mar 2026"
date-modified: "last-modified"
format: html
editor: visual
freeze: true
execute:
  eval: true
  echo: true
  warning: false
code-fold: true
---

# Bullet Charts, Sparklines and Dashboard Design

## 1 Overview

In this hands-on exercise, I learned how to create information dashboards using R.

By the end of this hands-on exercise, I was able to:

-   create bullet charts by using ggplot2,
-   create sparklines by using ggplot2,
-   build static information dashboards using gt and gtExtras, and
-   build interactive information dashboards using reactable and reactablefmtr.

## 2 Getting Started

### 2.1 Installing and launching R packages

I loaded the following R packages for this exercise.

The code chunk:

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse)
```

-   [tidyverse](https://www.tidyverse.org/) provides a collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. It is not a single package but a collection of modern R packages including but not limited to [readr](https://readr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [ggplot](https://ggplot2.tidyverse.org/), [tibble](https://tibble.tidyverse.org/), [stringr](https://stringr.tidyverse.org/), [forcats](https://forcats.tidyverse.org/) and [purrr](https://purrr.tidyverse.org/).
-   [lubridate]() provides functions to work with dates and times more efficiently.
-   [ggthemes]() is an extension of ggplot2. It provides additional [themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) beyond the basic themes of ggplot2.
-   [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) provides some additional helper functions to assist in creating beautiful tables with [gt](https://gt.rstudio.com/), an R package specially designed for anyone to make wonderful-looking tables using the R programming language.
-   [reactable](https://glin.github.io/reactable/index.html) provides functions to create interactive data tables for R, based on the [React Table](https://github.com/tanstack/table/tree/v7) library and made with [reactR](https://github.com/react-R/reactR).
-   [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) provides various features to streamline and enhance the styling of interactive reactable tables with easy-to-use and highly-customizable functions and themes.

## 3 Importing Microsoft Access Database

### 3.1 The Data Set

The data used in this exercise is a personal database in Microsoft Access mdb format called *Coffee Chain*.

### 3.2 Importing the Database into R

I used `odbcConnectAccess2007()` of [**RODBC**](https://cran.r-project.org/web/packages/RODBC/index.html) package to import the database query table into R. As this function requires a 32-bit ODBC driver, the code chunk below was run separately and the result saved as an rds file for subsequent use.

```{r}
#| eval: false
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

### 3.3 Data Preparation

I imported the pre-saved *CoffeeChain.rds* into R using the code chunk below.

```{r}
coffeechain <- read_rds("data/CoffeeChain.rds")
```

```{r echo=FALSE, eval=FALSE}
write_csv(coffeechain, "data/coffeechain.csv")
```

Next, I aggregated Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

### 3.4 Bullet Chart in ggplot2

I plotted bullet charts using ggplot2 functions to compare actual sales against targets for each product.

```{r}
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current),
           width=0.35,
           fill = "black") +
  geom_errorbar(aes(y = target,
                    x = Product,
                    ymin = target,
                    ymax= target),
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

::: callout-tip
## Things learned from the code chunk above

-   Multiple `geom_col()` layers with different fills create the background bands representing performance thresholds (100%, 75%, 50% of target).
-   `geom_errorbar()` with `ymin = ymax = target` draws a single horizontal line to mark the target value.
-   `coord_flip()` rotates the chart to make product labels easier to read.
:::

## 4 Plotting Sparklines using ggplot2

### 4.1 Preparing the Data

I prepared the data by filtering for 2013 records and computing monthly sales by product.

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

I then computed the minimum, maximum and end-of-month sales.

```{r}
mins <- group_by(sales_report, Product) %>%
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>%
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>%
  filter(Month == max(Month))

```

I also computed the 25th and 75th quantiles for each product.

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales,
                              0.25),
            quart2 = quantile(Sales,
                              0.75)) %>%
  right_join(sales_report)

```

### 4.2 Sparklines in ggplot2

I plotted sparklines using `ggplot2` with faceting by product, showing min/max points and the interquartile ribbon.

```{r}
#| fig-width: 14
#| fig-height: 12
ggplot(sales_report, aes(x=Month, y=Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2),
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank(), strip.text = element_blank())
```

::: callout-tip
## Things learned from the code chunk above

-   `facet_grid(Product ~ ., scales = "free_y")` creates a separate sparkline panel per product with independent y-axis scales.
-   `geom_ribbon()` uses the quartile data to shade the interquartile range as a reference band.
-   `geom_point()` highlights the minimum (red) and maximum (blue) sales months.
-   `theme_tufte()` from **ggthemes** gives the minimalist sparkline look with no gridlines or axes.
:::

## 5 Static Information Dashboard Design: gt and gtExtras Methods

In this section, I learned how to create a static information dashboard using [gt](https://gt.rstudio.com/index.html) and [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) packages. Before getting started, I reviewed the documentation of both packages to understand their key functions.

### 5.1 Plotting a Simple Bullet Chart

I used `gt_plt_bullet()` of **gtExtras** to create a bullet chart table, styled with `gt_theme_538()`.

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current,
              target = target,
              width = 60,
              palette = c("lightblue",
                          "black")) %>%
  gt_theme_538()
```

## 6 Sparklines: gtExtras Method

### 6.1 Preparing the Data

I prepared the sales report by filtering for 2013 data and summarising monthly sales by product.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date,
                        label = TRUE,
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()

```

I noted that gtExtras functions require list columns in the data frame. I used the code chunk below to convert the data into the required format.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")
```

### 6.2 Plotting the CoffeeChain Sales Report

```{r eval=FALSE}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

### 6.3 Adding Statistics

I computed summary statistics (min, max, average) for each product and displayed them in a formatted table.

```{r}
report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### 6.4 Combining the Data Frames

I joined the sparklines list column with the summary statistics to create a combined data frame.

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")

```

```{r}
sales <- report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

```{r}
sales_data = left_join(sales, spark)
```

### 6.5 Plotting the Combined Table

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

### 6.6 Combining Bullet Chart and Sparklines

I further combined the bullet chart with the sparklines to create a richer dashboard table.

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup()
```

```{r}
sales_data = sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue",
                          "black")) %>%
  gt_theme_538()
```

## 7 Interactive Information Dashboard Design: reactable and reactablefmtr Methods

In this section, I learned how to create an interactive information dashboard using [reactable](https://glin.github.io/reactable/index.html) and [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) packages. Before getting started, I reviewed the documentation of both packages to understand their key functions.

To build interactive sparklines, I first installed the **dataui** package from GitHub.

```{r}
#| eval: false
remotes::install_github("timelyportfolio/dataui")
```

I then loaded the package into the R environment.

```{r}
library(dataui)
```

### 7.1 Plotting Interactive Sparklines

Similar to the gtExtras approach, I first prepared the list column from the report data frame.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

I then used [react_sparkline](https://kcuilla.github.io/reactablefmtr/reference/react_sparkline.html) to plot the interactive sparklines.

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### 7.2 Changing the Page Size

By default the page size is 10. I used the `defaultPageSize` argument to display all 13 products on a single page.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### 7.3 Adding Points and Labels

I used `highlight_points` to mark the minimum and maximum values, and `labels` to annotate the first and last data points.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

### 7.4 Adding Reference Line

I used the `statline` argument to overlay a mean reference line on each sparkline.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

### 7.5 Adding Bandline

Instead of a reference line, I used the `bandline` argument to shade the inner quartile range as a band.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

### 7.6 Changing from Sparkline to Sparkbar

Finally, I switched from sparklines to sparkbars using `react_sparkbar()` to display the monthly sales values as bar charts.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```
