---
title: "Take-home Exercise 2: Exploratory & Confirmatory Analysis Module Prototype"
subtitle: "COFINFAD Analytics – Democratizing Fintech Intelligence"
author: "Lin Yan"
date: "March 14, 2026"
date-modified: last-modified
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: flatly
execute:
  warning: false
  message: false
  freeze: true
---

# Overview

## Background and Motivation

The **COFINFAD** (Colombia's Fintech Financial Advisory Dashboard) project aims to democratize fintech analytics for Colombia's financial ecosystem. The dataset comprises **48,723 customers** and over **3.16 million transactions** spanning a 12-month observation period, with variables capturing demographics, transaction behavior, product usage, satisfaction, and churn probability.

Exploratory and confirmatory analysis forms the first lens of the analytical narrative: it answers *what is happening* in the data before we ask *why* (segmentation) or *what will happen* (churn prediction). For banking practitioners, having a clear view of seasonal transaction patterns, how behavior differs by income or segment, and where customers and risk sit geographically supports both operational decisions and strategy.

## Module Objective

This prototype focuses on the **Exploratory & Confirmatory Analysis (EDA & CDA)** module of the COFINFAD Shiny application. The module enables analysts to:

-   **Explore** seasonal transaction volume by type (Transfer, Withdrawal, Payment, Deposit) via a time-series line chart
-   **Compare** customer demographics and transaction behavior across income brackets or segments with distribution and boxplots
-   **Test** whether group differences are statistically and practically significant (One-Way ANOVA or Welch's t-test with effect sizes η² and Cohen's d)
-   **Map** customer and risk distribution across Colombian departments (choropleth or bar chart)
-   **Inspect** correlations among key variables (tenure, satisfaction, churn, transaction metrics) via a heatmap

## Scope of This Report

This document serves as a prototype report following the structure recommended by the course. It covers:

1.  **Data preparation** — cleaning, imputation, feature engineering, transaction aggregation, and department-level statistics using only the two original CSVs
2.  **Visualization techniques** — working R code producing each chart and statistical test in the module
3.  **Design principles** — color encoding, interactivity, and layout rationale grounded in visual analytics practice
4.  **UI design** — layout, Shiny component mapping, and parameter/output specification

::: callout-note
## Packages Used

All R packages used in this prototype are available on **CRAN** and have been verified to work with R 4.3+:

| Package             | Purpose                                          |
|---------------------|--------------------------------------------------|
| `shiny`, `bslib`    | App framework and layout                         |
| `dplyr`, `tidyr`    | Data manipulation and aggregation                |
| `ggplot2`, `plotly` | Charts (static then interactive)                 |
| `scales`            | Axis labels (billions COP, percent)              |
| `leaflet`, `sf`     | Choropleth map (optional; fallback is bar chart) |
| `lubridate`         | Date handling (e.g. `floor_date` for month)      |
| `knitr`             | Table rendering                                  |

When `sf` or spatial data is missing, the app shows a horizontal bar chart by department instead of the map, so the module still runs. For the choropleth to be built, `ne_states(country = "colombia")` also requires the optional package **rnaturalearthhires** (e.g. `install.packages("rnaturalearthhires", repos = "https://ropensci.r-universe.dev", type = "source")`); otherwise Section 2.7 sets `dept_sf <- NULL` and Section 3.3 uses the bar fallback.
:::

```{r}
#| label: pkg-check
pkgs <- c(
  "shiny", "bslib", "dplyr", "tidyr", "ggplot2", "plotly",
  "scales", "lubridate", "knitr", "leaflet", "sf", "rnaturalearth"
)
for (p in pkgs) {
  ok <- requireNamespace(p, quietly = TRUE)
  cat(p, if (ok) "OK" else "MISSING", "\n")
}
```

# Data Preparation

## Data Sources

The raw data consists of two CSV files from the COFINFAD project:

| File | Records | Description |
|----|----|----|
| `customer_data (1).csv` | 48,723 rows | Demographics, product holdings, satisfaction scores, churn probability, location |
| `transactions_data (1).csv` | \~3.16M rows | Individual transactions with date, amount, type (Transfer/Withdrawal/Payment/Deposit), customer_id |

::: callout-important
## Data Quality Considerations

The raw data contains missing values in `credit_utilization_ratio`. Date and location fields require parsing and splitting for temporal and geographic analysis. These are addressed in the cleaning and imputation steps below.
:::

## Cleaning and Type Conversion

We load the two raw CSVs and clean the customer table so that later steps and the EDA have the right types and structure.

**Key decisions:**

-   **Date parsing**: `first_tx` and `last_tx` are converted to R `Date` type so we can compute recency and tenure
-   **Boolean conversion**: Product-holding columns stored as "True"/"False" are converted to logical for filtering and grouping
-   **Location splitting**: The composite `location` field is split into `city` and `department` for geographic analysis and the choropleth/bar chart
-   **Age grouping**: Continuous age is binned into 6 brackets for demographic comparison in distribution and boxplot views
-   **Factor levels**: `churn_risk`, `income_bracket`, and `customer_segment` are set with explicit levels so the order in charts and statistical comparisons is consistent

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(scales)
library(lubridate)
library(knitr)
library(sf)
library(rnaturalearth)
library(leaflet)
```

## Cleaning customer data

```{r}
#| label: load-raw
customers_raw <- read.csv("data/customer_data (1).csv", stringsAsFactors = FALSE)
transactions_raw <- read.csv("data/transactions_data (1).csv", stringsAsFactors = FALSE)
cat("Customers:    ", nrow(customers_raw), " rows, ", ncol(customers_raw), " cols\n")
cat("Transactions: ", nrow(transactions_raw), " rows, ", ncol(transactions_raw), " cols\n")
```

```{r}
#| label: clean-customers-2.1
customers <- customers_raw %>%
  mutate(
    first_tx = as.Date(first_tx),
    last_tx = as.Date(last_tx),
    savings_account = (savings_account == "True"),
    credit_card = (credit_card == "True"),
    city = trimws(sub(",.*", "", location)),
    department = trimws(sub(".*,", "", location)),
    age_group = cut(age, breaks = c(17, 25, 35, 45, 55, 65, 100),
                    labels = c("18-25", "26-35", "36-45", "46-55", "56-65", "65+")),
    churn_risk = factor(
      case_when(churn_probability < 0.2 ~ "Low",
                churn_probability < 0.35 ~ "Medium",
                TRUE ~ "High"),
      levels = c("Low", "Medium", "High")),
    income_bracket = factor(income_bracket, levels = c("Low", "Medium", "High", "Very High")),
    customer_segment = factor(customer_segment, levels = c("inactive", "occasional", "regular", "power"))
  )
```

## Missing Value Treatment

The most critical missing value issue is `credit_utilization_ratio`. We use **grouped median imputation** stratified by `income_bracket`:

::: callout-tip
## Why Grouped Imputation?

Credit use often varies by income — for example, low-income customers may have higher utilization ratios on average. If we filled missings with one overall median, we would flatten that structure and the distributions and tests by income bracket would no longer reflect real differences. Imputing with the median *within* each `income_bracket` keeps the relationship between income and credit behavior intact, so the EDA distributions and ANOVA/t-test results for this variable remain meaningful.
:::

```{r}
#| label: impute-2.2
customers <- customers %>%
  group_by(income_bracket) %>%
  mutate(credit_utilization_ratio = ifelse(
    is.na(credit_utilization_ratio),
    median(credit_utilization_ratio, na.rm = TRUE),
    credit_utilization_ratio
  )) %>%
  ungroup()
cat("Missing credit_utilization_ratio after imputation:",
    sum(is.na(customers$credit_utilization_ratio)), "\n")
```

## Feature Engineering

We add derived variables that the EDA uses to describe engagement and behavior. `recency_days` is the number of days from the customer’s last transaction to a fixed reference date; it supports the “how recently did they transact” view and is useful when comparing groups. Having these features in the customer table lets the distribution and boxplot charts compare behavior meaningfully (e.g. transaction volume or recency by income or segment) without recomputing on the raw log each time.

```{r}
#| label: features-2.3
customers <- customers %>%
  mutate(
    recency_days = as.numeric(as.Date("2024-01-01") - last_tx)
  )
```

## Transaction aggregation

The transaction log has over 3 million rows. If the app had to load and filter it on every user action, the EDA tab would feel slow. We therefore aggregate once at preparation time. We parse `date` and build a `month` variable with `floor_date(..., "month")`. Aggregation by month and transaction type gives `monthly_summary` — a small table (e.g. 12 months × 4 types) that the time series chart can use to show seasonal patterns by type without touching the raw log. `monthly_total` gives one row per month for date-range logic. We also aggregate by customer into `customer_tx_agg` (total amount, average amount, transaction count) and merge these into the customer table so each customer has `total_tx_volume`, `avg_tx_value`, and `tx_count`. The boxplots and distribution views then compare these metrics across income or segment from a single customer-level table, and the EDA stays responsive.

```{r}
#| label: transactions-2.4
transactions <- transactions_raw %>%
  mutate(
    date = as.Date(date),
    month = floor_date(date, "month")
  )

monthly_summary <- transactions %>%
  group_by(month, type) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE),
            avg_amount = mean(amount, na.rm = TRUE),
            tx_count = n(), .groups = "drop")

monthly_total <- transactions %>%
  group_by(month) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE),
            avg_amount = mean(amount, na.rm = TRUE),
            tx_count = n(), .groups = "drop")

customer_tx_agg <- transactions %>%
  group_by(customer_id) %>%
  summarise(
    total_tx_volume = sum(amount, na.rm = TRUE),
    avg_tx_value = mean(amount, na.rm = TRUE),
    tx_count = n(),
    .groups = "drop"
  )

# Join using renamed columns to avoid clash with any existing tx_count/total_tx_volume/avg_tx_value in raw customer data
customers <- customers %>%
  left_join(
    customer_tx_agg %>% rename(tx_count_agg = tx_count, total_tx_volume_agg = total_tx_volume, avg_tx_value_agg = avg_tx_value),
    by = "customer_id"
  ) %>%
  mutate(
    tx_count = coalesce(tx_count_agg, 0L),
    total_tx_volume = coalesce(total_tx_volume_agg, 0),
    avg_tx_value = coalesce(avg_tx_value_agg, 0)
  ) %>%
  select(-tx_count_agg, -total_tx_volume_agg, -avg_tx_value_agg)
```

## Department-level and spatial data (Guide 2.7)

We compute department-level statistics (`dept_stats`) from the customer table, then build the spatial object `dept_sf` by joining Colombia department boundaries (from `rnaturalearth`) with `dept_stats`. The choropleth in Section 3.3 uses `dept_sf` when available.

```{r}
#| label: dept-stats-2.7
dept_stats <- customers %>%
  group_by(department) %>%
  summarise(
    customer_count = n(),
    avg_churn_prob = mean(churn_probability, na.rm = TRUE),
    avg_income = mean(as.numeric(income_bracket), na.rm = TRUE),
    avg_clv = mean(customer_lifetime_value, na.rm = TRUE),
    avg_satisfaction = mean(satisfaction_score, na.rm = TRUE),
    .groups = "drop"
  )
```

**Spatial object for choropleth** (we do not save RDS here). `ne_states()` requires the optional package `rnaturalearthhires` for Colombia; if missing, the chunk sets `dept_sf <- NULL` and Section 3.3 falls back to a bar chart. The same logic (including department name normalization for the join) is used in the Shiny app’s `data_prep.R` (Section 7) so the app and this report stay aligned.

```{r}
#| label: dept-spatial
dept_sf <- NULL
tryCatch({
  library(sf)
  library(rnaturalearth)
  col_sf <- ne_states(country = "colombia", returnclass = "sf")

  # Normalise map department names (match app data_prep.R)
  col_sf$dept_match <- col_sf$name
  col_sf$dept_match[grepl("Bogot|Capital", col_sf$name, ignore.case = TRUE)] <- "Bogotá"

  # Normalise data department names for join (e.g. CSV "Bogota" -> "Bogotá")
  dept_stats_join <- dept_stats %>%
    mutate(department = case_when(
      grepl("^Bogota$", department, ignore.case = TRUE) ~ "Bogotá",
      TRUE ~ department
    ))

  dept_sf <- col_sf %>%
    select(dept_match, geometry) %>%
    left_join(dept_stats_join, by = c("dept_match" = "department"))

  # Report any unmatched departments (data departments not on map)
  unmatched <- unique(dept_stats_join$department[!dept_stats_join$department %in% col_sf$dept_match])
  if (length(unmatched) > 0) {
    cat("  Unmatched departments:", paste(unmatched, collapse = ", "), "\n")
    cat("  Attempting fuzzy match...\n")
    for (um in unmatched) {
      best <- col_sf$dept_match[which.min(adist(um, col_sf$dept_match))]
      cat("    ", um, " -> ", best, "\n")
    }
  }
  cat("  dept_sf built (", nrow(dept_sf), " rows). Choropleth available.\n")
}, error = function(e) {
  cat("  Warning:", conditionMessage(e), "\n")
  cat("  For choropleth: install sf, rnaturalearth, and rnaturalearthhires\n")
})
```

## Loading and Inspecting Prepared Data

The steps above produce `customers` (with imputed and engineered features plus transaction summaries), `monthly_summary`, `monthly_total`, and `dept_stats`. The following chunks verify structure and show a sample of the monthly summary used by the time-series chart.

```{r}
#| label: data-structure
cat("Customers:       ", nrow(customers), " rows\n")
cat("Monthly summary: ", nrow(monthly_summary), " rows\n")
cat("Department stats:", nrow(dept_stats), " departments\n")
```

```{r}
#| label: tbl-monthly-sample
monthly_summary %>%
  head(12) %>%
  knitr::kable(digits = 0, caption = "Sample of monthly_summary (first 12 rows)")
```

------------------------------------------------------------------------

# Visualization Techniques

This section demonstrates each visualization used in the EDA & CDA module. All outputs use the prepared data from the previous section.

## Color Palette

We define a consistent color scheme for transaction types and demographic groupings used across the EDA charts:

```{r}
#| label: color-palette
type_colors <- c("Transfer" = "#f97316", "Withdrawal" = "#fb7185",
                 "Payment" = "#fbbf24", "Deposit" = "#a855f7")
income_colors <- c("Low" = "#fb7185", "Medium" = "#fbbf24",
                   "High" = "#f97316", "Very High" = "#a855f7")
```

The same palette is used in the Shiny app so that transaction type and income bracket are encoded consistently across the time series, distribution, and boxplot views.

## Time series line chart (eda_ts)

Seasonal transaction volume often differs by transaction type — for example, Transfers may spike at year-end while Deposits are steadier. A time-series line chart, filterable by type and date range, lets analysts see when activity concentrates and which channels drive volume. We use `monthly_summary` (aggregated by month and transaction type) so the chart stays fast. The y-axis is in billions COP (/ 1e9), and each type has a fixed colour (Transfer=orange, Withdrawal=pink, Payment=amber, Deposit=purple) for consistency with the rest of the dashboard.

```{r}
#| label: fig-eda-ts
#| fig-cap: "Transaction volume over time by type"

type_colors <- c("Transfer" = "#f97316", "Withdrawal" = "#fb7185",
                "Payment" = "#fbbf24", "Deposit" = "#a855f7")

eda_date_range <- c(min(monthly_total$month), max(monthly_total$month))
eda_tx_types <- c("Transfer", "Withdrawal", "Payment", "Deposit")

d_ts <- monthly_summary %>%
  filter(type %in% eda_tx_types,
         month >= eda_date_range[1],
         month <= eda_date_range[2])

p_ts <- ggplot(d_ts, aes(x = month, y = total_amount / 1e9, color = type, group = type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2, alpha = 0.9) +
  scale_color_manual(values = type_colors) +
  scale_y_continuous(labels = label_comma(suffix = "B")) +
  labs(x = NULL, y = "Volume (Billions COP)", color = NULL) +
  theme_minimal()

ggplotly(p_ts, tooltip = c("x", "y", "colour"))
```

::: callout-note
## Observation

Transfer consistently has the highest transaction volume across the year, followed by Payment, while Deposit and Withdrawal remain much lower. This suggests that customer activity on the platform is mainly driven by transfers and payments. Another clear pattern is that all transaction types increase around June and August, which may reflect seasonal financial activity or a temporary rise in platform usage. After August, the volumes decline slightly and become more stable again. Overall, the chart suggests that both transaction type and time of year play a role in shaping customer transaction behavior.
:::

## Choropleth map

Customer counts and churn risk often cluster by region. We show one metric by department (here `customer_count`) with a YlOrRd colour scale on a dark basemap: leaflet map with polygons and legend. When spatial data (`dept_sf`) was not built we fall back to a horizontal bar chart by department.

```{r}
#| label: fig-eda-map
#| fig-cap: "Customer count by department (choropleth or bar fallback)"

eda_map_metric <- "customer_count"
if (!is.null(dept_sf) && nrow(dept_sf) > 0) {
  library(leaflet)
  library(htmltools)
  vals <- dept_sf[[eda_map_metric]]
  pal <- leaflet::colorNumeric("YlOrRd", domain = vals, na.color = "#333333")
  labels <- sprintf(
    "<strong>%s</strong><br/>%s: %s<br/>Customers: %s",
    dept_sf$dept_match,
    gsub("_", " ", eda_map_metric), round(vals, 2),
    ifelse(is.na(dept_sf$customer_count), "N/A",
           formatC(dept_sf$customer_count, format = "d", big.mark = ","))
  ) %>% lapply(htmltools::HTML)

  m <- leaflet(dept_sf) %>%
    addProviderTiles(providers$CartoDB.DarkMatter) %>%
    addPolygons(
      fillColor = ~pal(vals), fillOpacity = 0.7,
      color = "#444444", weight = 1, opacity = 0.8,
      label = labels,
      highlightOptions = highlightOptions(
        weight = 2, color = "#f97316", fillOpacity = 0.9, bringToFront = TRUE)
    ) %>%
    addLegend("bottomright", pal = pal, values = vals,
              title = gsub("_", " ", eda_map_metric), opacity = 0.8)
  print(m)
} else {
  d_map <- dept_stats %>% arrange(desc(.data[[eda_map_metric]]))
  ggplot(d_map, aes(x = reorder(department, .data[[eda_map_metric]]),
                    y = .data[[eda_map_metric]], fill = .data[[eda_map_metric]])) +
    geom_col(alpha = 0.8, width = 0.7) +
    scale_fill_distiller(palette = "YlOrRd", guide = "none") +
    coord_flip() +
    labs(x = NULL, y = "Customer count", title = "Customer count by department (bar fallback)") +
    theme_minimal()
}
```

## Distribution bar (eda_dist)

This chart shows how many customers fall into each level of a grouping variable (e.g. income_bracket or customer_segment), so we can see the demographic mix at a glance. In the app the user selects the grouping via `eda_group_var`; the same variable is then used in the boxplot and statistical test for a consistent story.

```{r}
#| label: fig-eda-dist
#| fig-cap: "Customer count by income bracket"

eda_group_var <- "income_bracket"
d_dist <- customers %>%
  count(!!sym(eda_group_var)) %>%
  arrange(desc(n)) %>%
  mutate(pct = n / sum(n))

p_dist <- ggplot(d_dist, aes(x = reorder(!!sym(eda_group_var), n), y = n,
                             text = paste0(!!sym(eda_group_var), ": ", n,
                                           " (", round(pct * 100, 1), "%)"))) +
  geom_col(fill = "#f97316", alpha = 0.8, width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Customer Count") +
  theme_minimal()

ggplotly(p_dist, tooltip = "text")
```

## Boxplot with y-axis trimming (eda_box)

We want to test whether transaction behavior (e.g. average transaction value or volume) varies meaningfully across income brackets or acquisition channel, and to show box plots next to the test results so users see both the distribution and the statistics. The boxplot gives a direct view by group. We trim the y-axis to the IQR×3 range with `coord_cartesian(ylim = ...)` so that a few extreme outliers do not squash the boxes and hide the main pattern; quantiles are still computed on the full data, so we are not dropping or censoring points, unlike `scale_y_continuous(limits = ...)` which would exclude points and distort the picture.

```{r}
#| label: fig-eda-box
#| fig-cap: "Avg transaction value by income bracket (y-axis trimmed)"

eda_group_var <- "income_bracket"
eda_test_var <- "avg_tx_value"
income_colors <- c("Low" = "#fb7185", "Medium" = "#fbbf24",
                   "High" = "#f97316", "Very High" = "#a855f7")

d_box <- customers %>%
  select(g = !!sym(eda_group_var), v = !!sym(eda_test_var)) %>%
  drop_na()

q1 <- quantile(d_box$v, 0.25, na.rm = TRUE)
q3 <- quantile(d_box$v, 0.75, na.rm = TRUE)
iqr <- q3 - q1
y_lo <- max(min(d_box$v, na.rm = TRUE), q1 - 3 * iqr)
y_hi <- min(max(d_box$v, na.rm = TRUE), q3 + 3 * iqr)

p_box <- ggplot(d_box, aes(x = g, y = v, fill = g)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.5, width = 0.6) +
  coord_cartesian(ylim = c(y_lo, y_hi)) +
  scale_fill_manual(values = income_colors) +
  labs(x = NULL, y = "Avg transaction value") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))

ggplotly(p_box)
```

## Confirmatory analysis: One-Way ANOVA (eda_test)

Test results are shown beside the plot so users can judge both statistical and practical significance. For three or more groups we use One-Way ANOVA and report F, p-value, and η² (eta-squared). With nearly 49,000 customers, p \< 0.05 can occur even when the group difference is tiny, so we also show a short note when the test is significant but η² \< 0.06 (small effect). That helps analysts avoid over-interpreting a significant p-value when the practical difference is negligible.

```{r}
#| label: eda-test-anova
d_test <- customers %>%
  select(group = !!sym(eda_group_var), value = !!sym(eda_test_var)) %>%
  drop_na()
d_test$group <- as.factor(d_test$group)

mod <- aov(value ~ group, data = d_test)
s <- summary(mod)
pv <- s[[1]]$`Pr(>F)`[1]
fv <- s[[1]]$`F value`[1]
eta2 <- s[[1]]$`Sum Sq`[1] / sum(s[[1]]$`Sum Sq`)

effect_interp <- function(e) {
  ifelse(e < 0.01, "negligible",
         ifelse(e < 0.06, "small",
                ifelse(e < 0.14, "medium", "large")))
}

cat("ONE-WAY ANOVA\n")
cat("Variable:     ", eda_test_var, "\n")
cat("Grouping:     ", eda_group_var, " (", nlevels(d_test$group), " groups)\n")
cat("Sample size:  ", format(nrow(d_test), big.mark = ","), "\n")
cat("F-statistic:  ", round(fv, 3), "\n")
cat("p-value:      ", formatC(pv, format = "e", digits = 3), "\n")
cat("Effect size η²:", round(eta2, 3), " — ", effect_interp(eta2), "\n")
if (pv < 0.05 && eta2 < 0.06)
  cat("Note: statistically significant but effect size is small (large n).\n")
```

## Welch's t-test (pairwise comparison)

For pairwise comparison (e.g. two acquisition channels or two income brackets) we use Welch's t-test (`var.equal = FALSE`) so we do not assume equal variances. We report Cohen's d as the effect size. In the app the user selects the two groups via dropdowns (`eda_group1`, `eda_group2`). Here we compare the first two levels of the grouping variable: **grouping variable** is `r eda_group_var`, and the two groups compared are **`r levels(d_test$group)[1]`** (Group 1) and **`r levels(d_test$group)[2]`** (Group 2). The output below prints these names again so the comparison is clear.

```{r}
#| label: eda-test-welch
eda_group1 <- levels(d_test$group)[1]
eda_group2 <- levels(d_test$group)[2]
if (nlevels(d_test$group) >= 2) {
  d_two <- d_test %>% filter(group %in% c(eda_group1, eda_group2))
  x1 <- d_two$value[d_two$group == eda_group1]
  x2 <- d_two$value[d_two$group == eda_group2]
  wt <- t.test(x1, x2, var.equal = FALSE)
  n1 <- length(x1); n2 <- length(x2)
  s1 <- sd(x1); s2 <- sd(x2)
  pooled_sd <- sqrt(((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2))
  cohens_d <- (mean(x1) - mean(x2)) / pooled_sd
  d_interp <- ifelse(abs(cohens_d) < 0.2, "negligible",
                    ifelse(abs(cohens_d) < 0.5, "small",
                           ifelse(abs(cohens_d) < 0.8, "medium", "large")))
  cat("WELCH'S T-TEST (two independent samples)\n")
  cat("Variable:     ", eda_test_var, "\n")
  cat("Group 1:      ", eda_group1, " (n = ", n1, ")\n")
  cat("Group 2:      ", eda_group2, " (n = ", n2, ")\n")
  cat("t-statistic:  ", round(wt$statistic, 3), "\n")
  cat("p-value:      ", formatC(wt$p.value, format = "e", digits = 3), "\n")
  cat("Cohen's d:    ", round(cohens_d, 3), " — ", d_interp, "\n")
  if (wt$p.value < 0.05 && abs(cohens_d) < 0.2)
    cat("Note: statistically significant but effect size is negligible (large n).\n")
} else {
  cat("Welch's t-test requires at least 2 groups. Current grouping has", nlevels(d_test$group), "level(s).\n")
}
```

## Correlation heatmap (eda_corr)

The correlation heatmap lets users explore how key customer attributes (e.g. tenure, satisfaction, churn probability, transaction metrics) relate to each other. We use `cor(..., use = "pairwise.complete.obs")` so that a few missing values do not wipe out whole rows or columns and shrink the matrix. The colour scale runs from purple (–1) to orange (+1), and each tile shows the r value for quick reading.

```{r}
#| label: fig-eda-corr
#| fig-cap: "Correlation heatmap (12 variables, pairwise.complete.obs)"

corr_cols <- customers %>%
  select(age, tx_count, avg_tx_value, total_tx_volume,
         satisfaction_score, churn_probability,
         credit_utilization_ratio, customer_tenure,
         active_products, nps_score, weekend_transaction_ratio,
         support_tickets_count)
cor_matrix <- cor(corr_cols, use = "pairwise.complete.obs")
cor_labels <- c("Age", "Tx Count", "Avg Tx Val", "Total Vol",
                "Satisfaction", "Churn Prob", "Credit Util",
                "Tenure", "Active Prod", "NPS", "Wknd Ratio", "Support Tix")

cor_df <- expand.grid(V1 = cor_labels, V2 = cor_labels, stringsAsFactors = FALSE)
cor_df$r <- as.vector(cor_matrix)
cor_df$V1 <- factor(cor_df$V1, levels = cor_labels)
cor_df$V2 <- factor(cor_df$V2, levels = rev(cor_labels))

p_corr <- ggplot(cor_df, aes(x = V1, y = V2, fill = r,
                              text = paste0(V1, " vs ", V2, "\nr = ", round(r, 2)))) +
  geom_tile(color = "grey20", linewidth = 0.5) +
  geom_text(aes(label = round(r, 2)), size = 2.2) +
  scale_fill_gradient2(low = "#a855f7", mid = "grey30", high = "#f97316",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_corr, tooltip = "text")
```

# Design Principles and Interactivity

This section discusses the visual analytics principles that guide the EDA module's design.

## Color Encoding Rationale

The EDA module uses a consistent palette: transaction types (Transfer=orange, Withdrawal=pink, Payment=amber, Deposit=purple) and income brackets (Low to Very High in pink/amber/orange/purple). The same variable is always encoded the same way across charts so users do not have to relearn encodings when switching views.

## Interactivity Design

The module follows an overview-first, then filter-and-detail pattern: the time series and map/bar give an overview of trends and geography; sidebar filters (date range, transaction types, grouping variable, test variable) drive the distribution, boxplot, and statistical test so that several views update together. Plotly tooltips show exact values; zoom, pan, and legend toggle are available. Test results (F/t, p-value, effect size) are displayed beside the boxplot so users can judge both statistical and practical significance.

## Clarity and Effect Sizes

With nearly 49,000 customers, p \< 0.05 can occur even when the group difference is tiny. We report effect sizes (η² for ANOVA, Cohen's d for t-test) together with p-values and show a short note when the test is significant but the effect size is small. The time series y-axis is in billions COP for readability. Boxplot y-axis is trimmed to IQR×3 so that extreme outliers do not squash the main pattern.

## Dark Theme

The app uses the team's dark theme (e.g. background #080808, accent #f97316) so that charts and KPIs match the rest of the COFINFAD dashboard.

# UI Design

## UI Design Process

The UI design for the COFINFAD Shiny application followed a sketch-first prototyping approach. The EDA module is one tab in the main navigation, with a sidebar for filters and a main area for KPI cards, time series, map/bar, distribution, boxplot, test result, and correlation heatmap.

::: callout-tip
## Why Hand-drawn Wireframes?

Hand-drawn sketches allow rapid iteration on layout and content without the overhead of pixel-perfect tooling. Once the layout and information hierarchy are agreed, they map directly to Shiny components (sidebar + main area, cards, plotlyOutput, etc.).
:::

## Shiny Component Mapping

### Sidebar Controls (Input Parameters)

-   **eda_date_range**, **eda_tx_types** — filter the time series and the transaction volume KPI.
-   **eda_reset** — reset date range and transaction types to defaults.
-   **eda_map_metric** — which metric to show on the map or bar (e.g. customer_count, avg_churn_prob, avg_income, avg_clv, avg_satisfaction).
-   **eda_group_var** — grouping variable for distribution, boxplot, and statistical test.
-   **eda_test_var** — numeric variable for the boxplot and for ANOVA/t-test.
-   **eda_stat_test** — choice between One-Way ANOVA and Welch’s t-test.
-   **eda_group1**, **eda_group2** — the two groups to compare when the user selects the t-test (choices come from `eda_group_var`).

## Outputs

-   **KPI row**: Total customers (fixed), transaction volume (filtered by date and type), average churn rate, average satisfaction.
-   **eda_ts**: Time series line chart (plotly).
-   **eda_map**: Choropleth (leaflet) when spatial data exists.
-   **eda_map_bar**: Horizontal bar by department when spatial data is not used.
-   **eda_dist**: Distribution bar (plotly).
-   **eda_box**: Boxplot with y-axis trimming (plotly).
-   **eda_test**: ANOVA or t-test result (F/t, p-value, effect size, conclusion) as UI (e.g. HTML).
-   **eda_corr**: Correlation heatmap (plotly).
-   **eda_demo**: Demographics chart (e.g. income × segment) in the app; I did not re-run it here but it uses the same `customers` data.

------------------------------------------------------------------------

------------------------------------------------------------------------

## Layout Structure

-   The EDA module is one tab (e.g. “EDA & CDA”) in the main navigation. **Left sidebar**: date range, transaction types, reset button, map metric, grouping variable, test variable, test type, and (when t-test is selected) group 1 and group 2 dropdowns. **Main area**: a row of KPI cards at the top; then time series and map/bar; then distribution; then boxplot and test result; then correlation heatmap (and in the app, demographics). The layout can be implemented with `bslib::layout_sidebar()` and a responsive grid of cards.

## Components Summary

1.  **KPI row** — Four value boxes: Total Customers (fixed 48,723), Transaction Volume (filtered by date and type), Avg Churn Rate, Avg Satisfaction.

2.  **Time series** — `plotlyOutput("eda_ts")`, driven by date range and transaction-type filters.

3.  **Map or bar** — `leafletOutput("eda_map")` when department geometry is available; otherwise `plotlyOutput("eda_map_bar")`. The metric is chosen via `eda_map_metric`.

4.  **Distribution** — `plotlyOutput("eda_dist")`, grouped by the selected variable.

5.  **Boxplot and test** — `plotlyOutput("eda_box")` and `uiOutput("eda_test")`, both using the chosen grouping and numeric variable; the test panel also uses the test type and, for t-test, the two selected groups.

6.  **Correlation** — `plotlyOutput("eda_corr")` for the 12-variable heatmap.

# Conclusion

This prototype demonstrates that the EDA & CDA module is both **technically feasible** and **analytically valuable** using standard CRAN packages. Key outcomes:

1.  **Data pipeline validated** — Cleaning, grouped imputation, feature engineering, transaction aggregation, and department-level statistics run on the two original CSVs and produce the tables required by all EDA charts and tests.
2.  **All visualizations and tests render correctly** — Time series, department bar, distribution, boxplot, ANOVA/t-test with effect sizes, and correlation heatmap produce interpretable output from the prepared data.
3.  **Design principles applied** — Consistent color encoding, overview-first interactivity, effect sizes alongside p-values, and dark theme alignment with the COFINFAD dashboard.
4.  **UI components mapped** — Every Shiny widget and output is specified with input IDs, purposes, and layout so the prototype can be implemented directly in the full application.
