---
title: "Hands-on_Ex8.1"
author: "Lin Yan"
date: "5 Mar 2026"
date-modified: "last-modified"
format: html
editor: visual
freeze: true
execute:
  eval: true
  echo: true
  warning: false
code-fold: true
---

# Choropleth Mapping with R

## 1 Overview

Choropleth maps symbolise areas (e.g. countries, subzones) with colours or patterns by value. This exercise uses [**tmap**](https://r-tmap.github.io/tmap/) to plot choropleth maps of Singapore (e.g. aged population by subzone).

## 2 Getting Started

Packages: [**tmap**](https://cran.r-project.org/web/packages/tmap/), [**readr**](https://readr.tidyverse.org/), [**tidyr**](https://tidyr.tidyverse.org/), [**dplyr**](https://dplyr.tidyverse.org/), [**sf**](https://cran.r-project.org/web/packages/sf/index.html). The last three come with **tidyverse**. Install and load:

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

::: callout-note
Install **tidyverse** only; it includes readr, tidyr, and dplyr.
:::

## 3 Importing Data into R

### 3.1 The Data

Two datasets:

-   **Geospatial:** Master Plan 2014 Subzone Boundary `MP14_SUBZONE_WEB_PL` (shapefile), [data.gov.sg](https://data.gov.sg/).
-   **Attribute:** Singapore Residents by Planning Area/Subzone, Age, Sex, Dwelling 2011–2020 `respopagesextod2011to2020.csv`, [SingStat](https://www.singstat.gov.sg/). Use fields PA and SZ to join to the shapefile.

### 3.2 Importing Geospatial Data into R

Import the shapefile with **sf**’s *st_read()* into `mpsz`:

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

Inspect `mpsz`:

```{r}
mpsz
```

### 3.3 Importing Attribute Data into R

Import the CSV with *read_csv()*:

```{r, echo=TRUE, message=FALSE}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

### 3.4 Data Preparation

Build a 2020 table with: PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

-   **YOUNG:** 0–4 to 20–24
-   **ECONOMY ACTIVE:** 25–29 to 60–64
-   **AGED:** 65+
-   **TOTAL:** all ages
-   **DEPENDENCY:** (YOUNG + AGED) / ECONOMY ACTIVE

#### 3.4.1 Data wrangling

Use *pivot_wider()* (tidyr), *mutate()*, *filter()*, *group_by()*, *select()* (dplyr):

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`), .groups = "drop") %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

#### 3.4.2 Joining the attribute data and geospatial data

Convert PA and SZ to uppercase so they match SUBZONE_N and PLN_AREA_N in the shapefile.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Join by SUBZONE_N = SZ using *left_join()* (keep `mpsz` on the left so the result stays an sf object).

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

Note: Using `mpsz` as the left table keeps the output as a simple feature data frame.

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## 4 Choropleth Mapping Geospatial Data Using *tmap*

Two options: quick *qtm()* or layered tmap elements for more control.

### 4.1 Plotting a choropleth map quickly by using *qtm()*

*qtm()* gives a fast default choropleth:

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

*tmap_mode("plot")* = static; *"view"* = interactive. *fill* maps the variable (e.g. DEPENDENCY).

### 4.2 Creating a choropleth map by using *tmap*'s elements

For finer control, use tmap layer functions (*tm_shape*, *tm_polygons*, *tm_fill*, etc.):

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

#### 4.2.1 Drawing a base map

Structure: *tm_shape(data)* + layer (e.g. *tm_polygons()*, *tm_fill()*). Base map:

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### 4.2.2 Drawing a choropleth map using *tm_polygons()*

Pass the variable name to *tm_polygons()* to shade by that variable:

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

Defaults: binning "pretty", colours YlOrRd, NA = grey. See 4.3 (classification) and 4.4 (colours).

#### 4.2.3 Drawing a choropleth map using *tm_fill()* and \*tm_border()\*\*

*tm_polygons()* = *tm_fill()* + *tm_borders()*. Fill only:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

Add borders with *tm_borders()*:

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,  
             fill_alpha = 0.1)
```

*tm_borders()*: *alpha* (0–1), *col*, *lwd*, *lty*.

### 4.3 Data classification methods of **tmap**

Classification groups values into bins. **tmap** styles: *fixed*, *sd*, *equal*, *pretty* (default), *quantile*, *kmeans*, *hclust*, *bclust*, *fisher*, *jenks*. Set *style* in *tm_fill()* or *tm_polygons()*.

#### 4.3.1 Plotting choropleth maps with built-in classification methods

Example (jenks, 5 classes):

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Example (*equal*):

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Quantile tends to spread counts more evenly than equal.

##### 4.3.1a Different Classification Methods

prepare choropleth maps by using different classification methods supported by tmap and compare their differences.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5)) +
  tm_title("Quantile Classification") +
  tm_borders(fill_alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_title("Equal Interval Classification") +
  tm_borders(fill_alpha = 0.5)
```

##### 4.3.1b Different Numbers of Classes

Preparing choropleth maps by using similar classification method but with different numbers of classes (i.e. 2, 6, 10, 20). Compare the output maps The number of classes can also affect how spatial patterns are interpreted.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 2)) +
  tm_title("Quantile Classification (2 classes)") +
  tm_borders(fill_alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 10)) +
  tm_title("Quantile Classification (10 classes)") +
  tm_borders(fill_alpha = 0.5)
```

::: callout-note
Observation

Different classification methods produce different spatial patterns. Quantile classification distributes observations evenly across categories, while equal interval classification may produce uneven distributions if the data are skewed.

Changing the number of classes also affects the visual interpretation of the map. A smaller number of classes simplifies the map but may hide important variation, whereas a larger number of classes reveals more detail but can reduce readability.
:::

#### 4.3.2 Plotting choropleth map with custome break

Custom *breaks*: use n+1 values for n classes (min and max included). Check *summary()* first:

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

Example breaks (n+1 values for n classes). Use max ≥ data max (here 20) to avoid "higher than highest break" warning:

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
    fill.scale = tm_scale_intervals(breaks = c(0, 0.60, 0.70, 0.80, 0.90, 20))) +
  tm_borders(fill_alpha = 0.5)
```

### 4.4 Colour Scheme

Colours: user-defined or **RColorBrewer**. Set *values* in *tm_scale_intervals()*.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

Use *"-brewer.greens"* (minus prefix) to reverse the ramp.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

### 4.5 Map Layouts

Layout = title, legend, scale bar, compass, margins. *tm_legend()* controls legend placement and format.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

#### 4.5.2 Map style

*tmap_style()* changes overall look. Example: *classic*.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("classic")
```

#### 4.5.3 Cartographic Furniture

*tm_compass()*, *tm_scale_bar()*, *tm_grid()* add compass, scale bar, and grid.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

Reset style: *tmap_style("white")*.

```{r}
tmap_style("white")
```

### 4.6 Drawing Small Multiple Choropleth Maps

Small multiples = several maps side by side (e.g. by variable or region). Options: (1) pass multiple variables to *fill*, (2) *tm_facets(by=)*, (3) *tmap_arrange()*.

#### 4.6.1 By assigning multiple values to at least one of the aesthetic arguments

Pass a vector of variable names to *tm_fill()* or *tm_polygons()*:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          fill.scale = tm_scale_intervals(style = "equal", values = "brewer.blues")) +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("white")
```

Multiple variables with different *style* and *palette*:

```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
    fill.scale = list(
      tm_scale_intervals(style = "equal", values = "brewer.blues"),
      tm_scale_intervals(style = "quantile", values = "brewer.greens"))) +
  tm_layout(legend.position = c("right", "bottom"))
```

#### 4.6.2 By defining a group-by variable in *tm_facets()*

*tm_facets(by = "REGION_N")*:

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.blues")) + 
  tm_facets(by = "REGION_N", 
            free.coords = TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(fill_alpha = 0.5)
```

#### 4.6.3 By creating multiple stand-alone maps with `tmap_arrange()`

Build separate map objects, then *tmap_arrange()*:

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.blues"))

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.blues"))

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### 4.7 Mappping Spatial Object Meeting a Selection Criterion

To map only a subset (e.g. one region), subset the sf object first:

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.blues"), 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(fill_alpha = 0.5)
```

## 5 Reference

### 5.1 All about **tmap** package

-   [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)
-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)
-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
-   [tmap: changes in version 2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)
-   [tmap: creating thematic maps in a flexible way (useR!2015)](http://von-tijn.nl/tijn/research/presentations/tmap_user2015.pdf)
-   [Exploring and presenting maps with tmap (useR!2017)](http://von-tijn.nl/tijn/research/presentations/tmap_user2017.pdf)

### 5.2 Geospatial data wrangling

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)
-   [Simple Features for R: StandardizedSupport for Spatial Vector Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)
-   [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)

### 5.3 Data wrangling

-   [dplyr](https://dplyr.tidyverse.org/)
-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
-   [tidyr: Easily Tidy Data with 'spread()' and 'gather()' Functions](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf)
